Towards an ecology of vagueness

### 2017-11-11 book_chapter/paper.tex
Wrote some more on the results of the loose interaction model.

We performed 25 simulation runs of this model for each of three population scenarios: only one population with $\alpha = 0$ (for reference), two populations with $\alpha \in \{0, 0.05\}$, and three populations with $\alpha \in \{0, 0.05, 0.1\}$.
For each scenario, starting proportions $P(\alpha)$ were equal for each value of $\alpha$.
Given the observations by~\textcite{franke_vagueness_2017} that state space size and tolerance parameter $\beta$ do not result in important qualitative difference, we fixed these values at $n_S = 30$ and $\beta = 0.1$.
We also used a fixed uniform distribution for the priors and a message space with 2 messages.
Regarding the duration of the simulations, due to the mixing in of random strategies at each step in order to simulate new individuals being introduced into the population (birth rate was fixed at $\gamma = 0.05$), the convergence criteria is no longer applicable.
Therefore, all simulation runs were stopped after 200 iterations.

@@

We ran the same number of simulation trials under the same conditions as for the model with tight population interaction.
%\begin{sidewaysfigure}
\begin{figure}
  \centering
  \begin{subfigure}[]{\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/proportion-cases-0-005-weakest.png}
    \caption{Two populations scenario.}
    \label{fig:proportion-cases-two-loose-interaction}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/proportion-cases-0-005-01-weakest.png}
    \caption{Three populations scenario.}
    \label{fig:proportion-cases-three-loose-interaction}
  \end{subfigure}
  \caption{Evolution of population proportions through time for each simulation trial of the loose interaction model.}
  \label{fig:proportion-cases-loose-interaction}
\end{figure}
%\end{sidewaysfigure}
In Figure~\ref{fig:proportion-cases-loose-interaction} we plot the evolution of population proportions for all trials of the multi-population scenarios.
The first thing to observe is that proportions evolve faster than in the model with tight interaction.
Whereas in the latter no given population ever reached much more than 70\% proportion, in this model we see that many trials resulted in one population fully dominating the others.
In the two-population scenario, some population reached 99\% proportion in 24 out of 25 trials.
In the three-population scenario, this happened in 18 out of 25 trials.
Given that strategies no longer develop in close tandem with each other, differences in expected utility can become more salient resulting in some type reaching a more clear advantage than the other.

More interestingly for our investigation, some trials actually resulted in a population with some degree of imprecision dominating the others.
For the two-population scenario, $\alpha = 0.05$ fully reached 100\% proportion in 8 trials, a point from which it is technically impossible for the other population to recover.
In most cases the dominating population gains its ground from the start, but in 5 cases we see a temporary advantage of $\alpha = 0.05$ that is then lost to the other population.
In one interesting trial (85), $\alpha = 0.05$ reached 99.87\% only to then steadly lose ground to $\alpha = 0$.
For the three population scenario, $\alpha = 0.05$ fully reached 100\% proportion in only 1 trial and $\alpha = 0.1$ never reached more than 59\%.
In 5 other trials we see a positive trend for $\alpha = 0.05$, but given the reversals that sometimes occur (\emph{e.g.}~trial 171), one cannot be sure whether full dominance would be achieved if the simulation was left to run for longer time.

>>>>

We performed 25 simulation runs of this model for each of three population scenarios: only one population with $\alpha = 0$ (for reference), two populations with $\alpha \in \{0, 0.05\}$, and three populations with $\alpha \in \{0, 0.05, 0.1\}$.
For each scenario, starting proportions $P(\alpha)$ were equal for each value of $\alpha$.
Given the observations by~\textcite{franke_vagueness_2017} that state space size and tolerance parameter $\beta$ do not result in important qualitative difference, we fixed these values at $n_S = 30$ and $\beta = 0.1$.
We also used a fixed uniform distribution for the priors and a message space with 2 messages.
Each type started with its own randomly generated strategy.
Regarding the duration of the simulations, due to the mixing in of new individuals into the population (birth rate was fixed at $\gamma = 0.05$), the convergence criteria is no longer applicable because each strategy is randomly perturbed at each time step.
Therefore, all simulation runs were stopped after 200 iterations.

@@

We ran the same number of simulation trials under the same conditions as for the model with tight population interaction.
%\begin{sidewaysfigure}
\begin{figure}
  \centering
  \begin{subfigure}[]{\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/proportion-cases-0-005-weakest.png}
    \caption{Two populations scenario.}
    \label{fig:proportion-cases-two-loose-interaction}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/proportion-cases-0-005-01-weakest.png}
    \caption{Three populations scenario.}
    \label{fig:proportion-cases-three-loose-interaction}
  \end{subfigure}
  \caption{Evolution of population proportions through time for each simulation trial of the loose interaction model.}
  \label{fig:proportion-cases-loose-interaction}
\end{figure}
%\end{sidewaysfigure}
In Figure~\ref{fig:proportion-cases-loose-interaction} we plot the evolution of population proportions for all trials of the multi-population scenarios.
The first thing to observe is that proportions evolve faster than in the model with tight interaction.
Whereas in the latter no given population ever reached much more than 70\% proportion, in this model we see that many trials resulted in one population fully dominating the others.
In the two-population scenario, some population reached 99\% proportion in 24 out of 25 trials.
In the three-population scenario, this happened in 18 out of 25 trials.
%Given that strategies no longer develop in close tandem with each other, differences in expected utility can become more salient resulting in some type reaching a more clear advantage than the other.
More interestingly for our investigation, some trials actually resulted in a population with some degree of imprecision dominating the others.
For the two-population scenario, $\alpha = 0.05$ fully reached 100\% proportion in 8 trials (79, 86, 88, 89, 91, 93, 96, 98), a point from which it is technically impossible for the other population to recover.
In most cases the dominating population gains its ground from the start, but in 5 cases we see a temporary advantage of $\alpha = 0.05$ that is then lost to the other population.
In one interesting trial (85), $\alpha = 0.05$ reached 99.87\% only to then steadly lose ground to $\alpha = 0$.
For the three population scenario, $\alpha = 0$ was reduced to 0\% proportion in 4 trials (152, 157, 159, and 175).
In all of those cases, $\alpha = 0.05$ clearly has the upper hand over $\alpha = 0.1$, despite a temporary advantage of the latter in some trials.
%In 5 other trials we see a positive trend for $\alpha = 0.05$, but given the reversals that sometimes occur (\emph{e.g.}~trial 171), one cannot be sure whether full dominance would be achieved if the simulation was left to run for longer time.

Because of the loose interaction between populations, different types can now evolve separately.
One consequence of this is that populations would a certain level of imprecision again reach convexity faster than those without imprecision.
\begin{figure}
  \centering
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/convex-cases-0-005-weakest.png}
    \caption{Two populations scenario.}
    \label{fig:convex-cases-two-loose-interaction}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/convex-cases-0-005-01-weakest.png}
    \caption{Three populations scenario.}
    \label{fig:convex-cases-three-loose-interaction}
  \end{subfigure}
  \caption{First iteration with convex sender strategies for each simulation trial of the loose interaction model.}
  \label{fig:convex-cases-loose-interaction}
\end{figure}
In Figure~\ref{fig:convex-cases-loose-interaction} we plot, for each trial, the first iteration when each type reached convexity.
What we see is that, even though that is true on average, it is not always the case.
This has certainly to do with the initial conditions of each trial, since the randomly generated strategies can simply by chance be more favourable to reaching convexity.
More importantly, we also see that reaching convexity sooner is not a suficient condition for achieving population dominance .
There were many trials where another type reached convexity much sooner than $\alpha = 0$ but was completely dominated nevertheless (\emph{e.g.}~76, 84, 92, and 99 in the two-population scenario; 156, 165, 168, and 172 in the three-population scenario).
It also does not seem to be fully necessary, given a few examples where $\alpha = 0$ reached convexity early on and another type ended up dominating (89 in the two-population scenario and 152 in the three-population scenario).

Another consequence of the two populations evolving separately is that they do not necessarily evolve towards the same equilibrium.
In a sim-max game such as the one set up there are only two stable equilibria.
These are two Voronoi systems of the kind shown in Figure~\ref{fig:example_strats}, one where the first message is used for the first half of the state space, and the other where the second message is used for this region.
In the multi-population model with loose interaction, each population evolves independently towards one of these two equilibria.
This is importantly conditioned by the initial strategy the process of selection initiates from.
The populations are, however, not fully independent, since the process of selection of the level of precision relies on the expected utility of one population playing against itself and the others.
And this is important because strategies in one equilibrium get the lowest payoff possible against strategies in the other equilibrium.
Now, if two populations evolve towards different equilibria, whatever advantage one population has playing against itself could trigger a runaway effect by causing an increase in its proportion, which in turn will increase the population's relative expected utility, potentially increasing their proportion further in the next round, and so forth.
In this case, one would expect that faster convergence towards convexity would be especially important for success.

In the two-population scenario, of the 8 trials where $\alpha = 0$ was reduced to 0\%, all but two (88, 98) ended with the two populations each close to a different equilibria.
In the two-population scenario, the population with $\alpha = 0$ evolved towards the same equilibrium as the other two in 1 of the trials where it was eliminated.
In the other 3 trials, in 2 it evolved to a different equilibrium than the other two, and in 1 it evolved towards the same equilibrium as $\alpha = 0.05$ (but different to the one $\alpha = 0.1$ evolved towards).
Again, we do not find a clear case for this being a decisive factor in the success of populations with impairment.
That same goes for the impact of an initial advantage in expected utility creating the runaway effect we just mentioned.
Despite none of these three factors (including reaching convexity sooner) sentencing the demise of $\alpha = 0$ with certainty, they do seem to conspire together to bring it about.
In most trials where the type was eliminated, in both the two- and three-population scenarios, $\alpha = 0$ ended up evolving towards a different equilibrium than the other types, and either had an initial disadvantage in expected utility, or reached convexity much later.
These factors can thus be seen as indicators at best, but the story of how a population with imprecision ends up dominating one without seems to be more complicated to tell.

This is not surprising since we are facing a complex dynamical system with various interacting components (sender and receiver strategies and multiple populations).
\begin{figure}
  \centering
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/sender-eu-trial-88-0-005-weakest.png}
    \caption{Sender expected utility.}
    \label{fig:sender-eu-trial-88-loose-interaction}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{simulation/results/round-3/plots/receiver-eu-trial-88-0-005-weakest.png}
    \caption{Receiver expected utility.}
    \label{fig:receiver-eu-trial-88-loose-interaction}
  \end{subfigure}
  \caption{Expected utilities for trial 88 of the loose interaction model. Vertical lines demarcate, for each type, the iteration where convexity was achieved.}
  \label{fig:eu-trial-88-loose-interaction}
\end{figure}
Just as an example of this, in Figure~\ref{fig:eu-trial-88-loose-interaction} we plot the expected utilities of sender and receiver strategies for both populations of trial 88.
The curves up to where $\alpha = 0.05$ achieved convexity seem to suggest that the sender strategy of that type was evolving in tandem with the receiver strategy of the other.
This is possible in our model, although we expect it to be highly unlikely.
The moment where the population reaches convexity marks the point where the sender strategy of $\alpha = 0.05$ aligns with its type's receiver strategy and this coincides with the moment where the population seems to gain real traction over the other type (see again the plot for trial 88 in Figure~\ref{fig:convex-cases-loose-interaction}).
